---
title: "Milwaukee Bucks Analysis"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Question 1
# Situation

- 4th quarter
- Less than 24 seconds remaining in the game
- Shot clock is off
- Milwaukee Bucks are leading by 3 points
- Milwaukee Bucks are on defense
  
# Coaches asks whether we should foul?
-   make recommendation to the coach on a prescribed course of action
-   1-2 pages
-   understood by both technical and non technical
-   indicate what other data points (not provided) that would help improve your responses
  


```{r,warning=FALSE}

library(dplyr)
library(ggplot2)


```

```{r,warning=FALSE}

# Load the data
data <- read.csv("chances.csv")

# Data Quality check

## Check for missing values in a data frame

data[data == "null"] <- NA
sum(is.na(data))

## Explore the data structure
str(data)

## Correct the datatypes
data<-data%>%
  mutate(
    gid_nba = as.character(gid_nba),
    game_date = as.Date(game_date, format = "%m/%d/%Y"),
    shot_fga = as.integer(shot_fga),
    shot_fg = as.integer(shot_fg),
    shot_fga3 = as.integer(shot_fga3),
    shot_fg3 = as.integer(shot_fg3),
    shot_qSP = as.double(shot_qSP),
    fouls_def = as.integer(fouls_def),
    fouls_def_shooting = as.integer(fouls_def_shooting),
    fta = as.integer(fta),
    fta_shooting = as.integer(fta_shooting)
  )


## Check for duplicate rows in a data frame
any(duplicated(data))


```


```{r} 

# subset the data for the situation from question 1

filtered_data <- data %>%
  filter( # 4th quarter 
          period == 4,                       
          # Milwaukee bucks team on defense
          team_nba_def == "MIL",             
          # startType of chance is advantageous to defense team
          startType %in% c("FGDRB", "TO", "FTDRB", "SLOB", "BLOB"))  
        
head(filtered_data)

# Calculate the success rate when fouling (mean of shot_qSP)
filtered_data$shot_qSP[is.na(filtered_data$shot_qSP)] <- 0
success_rate_foul <- round(mean(as.numeric(filtered_data$shot_qSP)),2)

# Calculate the success rate when not fouling
success_rate_no_foul <- 100-success_rate_foul


# Create a data frame for the success rates
success_rates <- data.frame(
  Category = c("No Foul", "Foul"),
  Rate = c(success_rate_no_foul, success_rate_foul)
)

# Create a pie chart
ggplot(success_rates, aes(x = "", y = Rate, fill = Category)) +
  geom_bar(stat = "identity", width = 1) +
  geom_col(color = "black")+
  coord_polar(theta = "y") +
  labs(title = "Graph1: Success Rates of Fouling", fill = NULL) +  
  theme_void() +
  scale_fill_manual(values = c("No Foul" = "#00471B", "Foul" = "#EEE1C6")) +
  theme(legend.position = "right",  
        legend.text = element_text(size = 12), 
        plot.title = element_text(size = 16, hjust = 0.5),  
        
        plot.margin = unit(rep(1, 4), "cm")) +
  
  geom_text(aes(label = paste0(Rate, "%")), position = position_stack(vjust = 0.5), size = 5)


# Make a recommendation based on success rates
if (success_rate_foul > success_rate_no_foul) {
  recommendation <- "Based on historical data, fouling is recommended."
} else {
  recommendation <- "Based on historical data, not fouling is recommended."
}

# Recommendation
cat("Success Rate with Fouling:", success_rate_foul, "and Success Rate without Fouling:", success_rate_no_foul, "%\n","-> Hence:",recommendation, "\n")
```


```{r}

# find frequency of (startTyoe,outcome) pairs for our team

new_filtered_data <- data %>%
  filter( # 4th quarter 
          period == 4,                       
          # Milwaukee bucks team on defense
          team_nba_def == "MIL" | team_nba_off=="MIL",             
          # startType of chance is advantageous to defense team
          startType %in% c("FGDRB", "TO", "FTDRB", "SLOB", "BLOB")) 

# Create a table of (startType, outcome) pairs
pair_frequency <- table(new_filtered_data$startType, new_filtered_data$outcome)

# Convert the table to a data frame for better readability
pair_frequency_df <- as.data.frame(pair_frequency)

# Rename the columns for clarity
colnames(pair_frequency_df) <- c("startType", "outcome", "frequency")

# Print the pair frequency data frame
print(pair_frequency_df)

# bar chart for best kind of start type

# Filter out pairs with a frequency of 0
pair_frequency_filtered <- pair_frequency_df[pair_frequency_df$frequency > 0, ]

# Create a bar chart
barplot(
  height = pair_frequency_filtered$frequency,
  names.arg = paste(pair_frequency_filtered$startType,' X ', pair_frequency_filtered$outcome),
  main = "Graph3:Frequency of (startType, outcome) Pairs",
  ylab = "Frequency",
  col = "#00471B",            
  border = "black",           
  xlim = c(0.5, nrow(pair_frequency_filtered) + 0.5),  
  ylim = c(0, max(pair_frequency_filtered$frequency) + 5),
  las = 2,                    
  space = 0.2,                
  cex.names = 0.5             
)

```
```{r}

# Convert the table to a matrix
pair_frequency_matrix <- as.matrix(pair_frequency)

# Create a heatmap
ggplot(pair_frequency_df, aes(x = outcome, y = startType, fill = frequency)) +
  geom_tile() +                                          
  geom_text(aes(label = frequency), vjust = 1) +         
  scale_fill_gradient(low = "white", high = "#00471B") +  
  labs(title = "Graph2: Heatmap of (startType, outcome) Pair Frequencies",
       x = "Outcome", y = "Start Type") +                
  theme_minimal()+                                       
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
